Linear stability of magnetized massive protoplanetary disks

### 2014-02-10 paper/linear.tex
major update to tex

\section{Linear problem}\label{linear}
We consider axisymmetric Eulerian perturbations to the above
equilibrium in the form $\real[\dd\rho(z)\exp{\imgi(k_xx + \sigma
    t)}]$ and similarly for other fluid variables. Here, $k_x$ is a constant radial wavenumber and 
$\sigma=-(\omega +\imgi\gamma)$ is a complex 
frequency, where $-\omega$ is the real mode frequency and $\gamma$ is
the growth rate. Hereafter, we surpress the exponential factor, as
well as the real part notation.  We further write 
\begin{align}
  &z=\hat{z}H,\quad k_x =  k_x/H,\quad \sigma = \hat{\sigma}\Omega, \quad \delta\bm{v} = \csmid
  \delta\hat{\bm{v}}, \\ 
  &\delta\bm{B} = B\delta\hat{\bm{B}},\quad
  \delta\rho = \rho\hat{W}/\hat{c}_s^2,\quad \delta\Phi =
  \csmid^2\delta\hat{\Phi}, 
\end{align}
where we have introduced the non-dimensional enthalpy perturbation
$\hat{W}$ and the sound-speed $\hat{c}_s=\csmid^{-1}\sqrt{dP/d\rho}$ can be 
obtained from the equation of state. The background frequencies are
written $S=\hat{S}\Omega$ and $\Omega_z=\hat{\Omega}_z\Omega$ and the 
resistivity as $\eta = \hat{\eta}H^2\Omega$.

We will now drop the $\hat{\phantom{a}}$ notation. Henceforth it is
understood that all variables have been appropriately normalized. The
linearized continuity equation is 
\begin{align}
  \frac{\imgi\sigma}{fc_s^2}W + \imgi k_x \dd v_x
  +\left(\ln\rho\right)^\prime\dd v_z + \dd v_z^\prime = 0,\label{lin_cont}
\end{align}
where $^\prime$ denotes $d/dz$. The linearized equations of motion are
\begin{align}
  &\imgi\sigma \dvx - 2\dvy = f\left[v_A^2\dbx^\prime - \imgi k_x
    \left(\w + v_A^2\dbz
    \right)\right],\\
  &\imgi\sigma\dvy + \frac{\kappa^2}{2}\dvx = fv_A^2\dby^\prime,\\
  & \imgi\sigma\dvz = -f\w^\prime,
\end{align}
where the Alfven speed $v_A =
\left(\beta\rho\right)^{-1/2}$, the effective enthalpy perturbation
$\w = W + \dphi$ and the epicycle frequency $\kappa=\sqrt{2(2-S)}$. 
The linearized induction equation is
\begin{align}
&  \imgi\bar{\sigma}\dbx = f\dvx^\prime +
  \eta\dbx^{\prime\prime}+\eta^\prime\dbx^\prime - \imgi k_x
  \eta^\prime \dbz,\label{induct_x}\\
&\imgi\bar{\sigma}\dby = f\dvy^\prime - S\dbx +
  \eta\dby^{\prime\prime}+\eta^\prime\dby^\prime,\label{induct_y}\\
& \imgi\bar{\sigma}\dbz = -\imgi f k_x \dvx + \eta\dbz^{\prime\prime}, \label{induct_vert}
\end{align}
where $\imgi\bar{\sigma} = \imgi\sigma + \eta k_x^2$, and the
divergence-free condition is $\imgi k_x\dbx + \dbz^\prime=0$. Finally,
the linearized Poisson equation is
\begin{align}
  \dphi^{\prime\prime} - k_x^2\dphi = \frac{\rho}{c_s^2f^2Q}W.  \label{lin_poisson}
\end{align}

We now eliminate $\dd\bm{B}$ and $\dvz$ between the linearized
equations to obtain a system of ordinary differential equations for
$\bm{U}=\left(\dvx,\dvy,W,\dphi\right)$. We detail this technical
procedure in Appendix \ref{reduction}. The schematic problem is
\begin{align}
  L_{11}\dvx + L_{12}\dvy + L_{13}W + L_{14}\dphi &= 0, \label{lin1}\\
  L_{21}\dvx + L_{22}\dvy + L_{23}W + L_{24}\dphi &= 0, \label{lin2}\\
  L_{31}\dvx \quad\quad\quad\,\,\,\,\,\,\, + L_{33}W + L_{34}\dphi &=0,\label{lin3}\\
  \phantom{L_{31}\dvx + L_{22}\dvy +} L_{43}W + L_{44}\dphi
  &=0\label{lin4}, 
\end{align}
where the differential operators $L_{1j}$, $L_{2j}$ and $L_{3j}$ can
be read off Eq. \ref{final_vx}, Eq. \ref{final_vy} and 
Eq. \ref{final_w} respectively, and $L_{4j}$ from
Eq. \ref{lin_poisson}.

Note that $L_{31}\propto k_x$, so that if $k_x=0$ 
then Eq. \ref{lin1}---\ref{lin2} are decoupled from 
Eq. \ref{lin3}---\ref{lin4}.

\subsection{Boundary conditions}
We assume $\bm{U}$ is an even function of $z$, so that $d\bm{U}/dz=0$
at $z=0$. This implies that $\dbx=\dby=0$ at the midplane, consistent
with a highly resistive dead zone. At the upper disk boundary
$z=\zmax$ we assume 
$\dvz=\dbx=\dby=0$, and 
\begin{align}
  \dphi^\prime(\zmax) + k_x\dphi(\zmax) = 0
\end{align}
for the potential perturbation \citep[see][]{goldreich65a}.

\subsection{Numerical procedure}
We use a pseudo-spectral approach to solve the set of linearized
equations. Let
\begin{align}\label{cheby_expand}
  \bm{U}(z) 
  = \sum_{k=1}^{N_z} \bm{U}_k\psi_k(z/\zmax), 
\end{align}
where $  \psi_k  = T_{2(k-1)},  $ and  $T_l$ is a Chebyshev polynomial of the first
kind of order $l$ \citep{stegun65}. %The number of basis functions is
%$N_z$ and $l_\mathrm{max}=2(N_Z-1)$ is the highest polynomial order. 
Note that we only use even Chebyshev polynomials which automatically satisifies the symmetry 
condition at $z=0$. Henceforth we only consider $z\geq 0$. 
The pseudo-spectral coefficients $\bm{U}_n$  are obtained by demanding
the set of linear equations to be satisfied at $N_z$ collocation
points along the vertical direction. We choose these points as the
extrema of $T_{2(N_z-1)}$ plus end points.

This procedure discretizes the linear equations to a matrix equation,
\begin{align}\label{matrix_eqn}
\bm{M}\bm{w} = 0, 
\end{align}
where $\bm{M}$ is a $4N_z\times 4 N_z$ matrix representing the $L_{ij}$ 
and upper disk boundary conditions, 
and $\bm{w}$ is a vector storing the pseudo-spectral coefficients. 
Starting with an initial guess for $\sigma$, non-trivial solutions to
Eq. \ref{matrix_eqn} are obtained by varying $\sigma$ using Newton-Raphson iteration 
such that $\mathrm{det}\bm{M}=0$ \citep[details can be found in][]{lin12}.

\subsection{Diagnostics}

>>>>

\section{Linear problem}\label{linear}
We consider axisymmetric Eulerian perturbations to the above
equilibrium in the form $\real[\dd\rho(z)\exp{\imgi(k_xx + \sigma
    t)}]$ and similarly for other fluid variables. Here, $k_x$ is a
constant radial wavenumber and $\sigma=-(\omega +\imgi\gamma)$ is a
complex  frequency, where $-\omega$ is the real mode frequency and $\gamma$ is
the growth rate. We take $k_x>0$ without loss of generality. 
Hereafter, we surpress the exponential factor, as
well as the real part notation.

The linearized continuity equation is 
\begin{align}
  \frac{\imgi\sigma}{c_s^2}W + \imgi k_x \dd v_x
  +\left(\ln\rho\right)^\prime\dd v_z + \dd v_z^\prime = 0,\label{lin_cont}
\end{align}
where $^\prime$ denotes $d/dz$ and $W=\delta P/\rho =
c_s^2\delta\rho/\rho$ is the enthalpy perturbation.  
The linearized equations of motion are
\begin{align}
  &\imgi\sigma \dvx - 2\Omega\dvy  = - \imgi
  k_x\w + \frac{B_z}{\mu_0\rho} \left[\dbx^\prime - \imgi 
    k_x\left(\dbz + \epsilon\dby\right)\right],\\
  &\imgi\sigma\dvy + 
    \frac{\kappa^2}{2\Omega}\dvx = \frac{B_z}{\mu_0\rho}\dby^\prime\label{lin_vy},\\ 
  & \imgi\sigma\dvz = -\w^\prime - \frac{B_y}{\mu_0\rho}\dby^\prime\label{lin_vz},  
\end{align}
where the effective enthalpy perturbation
$\w = W + \dphi$. The linearized induction equation is
\begin{align}
&  \imgi\bar{\sigma}\dbx = B_z\dvx^\prime +
  \eta\dbx^{\prime\prime}+\eta^\prime\dbx^\prime - \imgi k_x
  \eta^\prime \dbz,\label{induct_x}\\
&\imgi\bar{\sigma}\dby = B_z\dvy^\prime -B_y\Delta - S\dbx +
  \eta\dby^{\prime\prime}+\eta^\prime\dby^\prime,\label{induct_y}\\
& \imgi\bar{\sigma}\dbz = -\imgi  k_xB_z \dvx +
\eta\dbz^{\prime\prime}, \label{induct_vert}
%& \imgi\bar{\sigma}\dbz = -\imgi  k_x \dvx  -\imgi k_x\eta \dbx^{\prime}, \label{induct_vert}
\end{align} 
where $\imgi\bar{\sigma} = \imgi\sigma + \eta k_x^2$,
$\Delta\equiv \nabla\cdot\bm{\delta v}  = \imgi k_x\dvx + \dvz^\prime$,   
and the divergence-free condition is $\imgi k_x\dbx +
\dbz^\prime=0$. Finally, the linearized Poisson equation is 
\begin{align}
  \dphi^{\prime\prime} - k_x^2\dphi = \frac{\Omega^2\rho}{c_s^2Q\rho_0}W.  \label{lin_poisson}
\end{align}

We eliminate $\dd\bm{B}$ and $\dvz$ between the linearized
equations to obtain a system of ordinary differential equations for
$\bm{U}=\left(\dvx,\dvy,W,\dphi\right)$. We detail the steps 
in Appendix \ref{reduction} for two cases considered in this paper:
%We consider two cases:
\begin{enumerate}
\item Purely vertical field with constant or variable
  resistivity (see below), so that $\epsilon = 0$ and $\eta=\eta(z)$. 
\item Tilted field with uniform resistivity
  so that $\epsilon \neq 0$ and $\eta=\rm{constant}$.   
\end{enumerate}
Schematically, the numerical problem is to solve 
\begin{align}
  L_{11}\dvx + L_{12}\dvy + L_{13}W + L_{14}\dphi &= 0, \label{lin1}\\
  L_{21}\dvx + L_{22}\dvy + L_{23}W + L_{24}\dphi &= 0, \label{lin2}\\
  L_{31}\dvx +L_{32}\dvy + L_{33}W + L_{34}\dphi &=0,\label{lin3}\\
  \phantom{L_{31}\dvx + L_{22}\dvy +} L_{43}W + L_{44}\dphi
  &=0\label{lin4},
\end{align}
where differential operators $L_{1j},\,L_{2j}$ and $L_{3j}$
($j=1,2,3,4$) can be read off Appendix \ref{reduction} and $L_{4j}$
($j=3,4$) from the linearized Poisson equation above. We remark that
the more general problem of tilted field and variable resistivity
%($\epsilon\neq0$ and $\eta=\eta(z)$
can also be reduced to the above form.

%where the differential operators $L_{1j}$, $L_{2j}$ and $L_{3j}$ can
%be read off Eq. \ref{final_vx}, Eq. \ref{final_vy} and 
%Eq. \ref{final_w} respectively, and $L_{4j}$ from
%Eq. \ref{lin_poisson}.

%Note that $L_{13},\,L_{14},\,L_{23},\,L_{24}$ and $L_{31}$ are
%proportional to $k_x$. Then for for $k_x=0$ 
%Eq. \ref{lin1}---\ref{lin2} are decoupled from 
%Eq. \ref{lin3}---\ref{lin4}. In this case, a solution is $\dphi=W=0$,
%with $\dvx$ and $\dvy$ determined by $L_{11}\dvx + L_{12}\dvy = 0$ and 
%$L_{21}\dvx + L_{22}\dvy=0$, which is the incompressible MRI problem. 
%This means that the density and potential perturbations can only
%influence the MRI for $k_x\neq0$. Nevertheless, for $k_x=0$,
%self-gravity can still have an effect through the background
%equilibrium.

\subsection{Boundary conditions}
At the vertical disk boundaries $z=\pm\zmax$ we assume  
$\dvz=\dbx=\dby=0$, and 
\begin{align}
  \dphi^\prime(\zmax) \pm k_x\dphi(\zmax) = 0
\end{align}
for the potential perturbation \citep[see][]{goldreich65a}, where the
upper (lower) sign applies to the upper (lower) disk surface. 
For Case 1 (purely vertical field) we additionally assume $\bm{U}$ is
an even function of $z$, so that $d\bm{U}/dz=0$ 
at $z=0$.

%This implies that $\dbx=\dby=0$ at the midplane, consistent
%with a highly resistive dead zone.

\subsection{Numerical procedure}
We use a pseudo-spectral method to solve the set of linearized
equations. Let
\begin{align}\label{cheby_expand}
  \bm{U}(z) 
  = \sum_{k=1}^{N_z} \bm{U}_k\psi_k(z/\zmax), 
\end{align}
where 
\begin{align}
  \psi_k  = 
  \begin{cases}
    T_{2(k-1)} & \text{Vertical field (case 1)},\\
    T_{k-1}   & \text{Tilted field (case 2)},
  \end{cases}
\end{align}
and  $T_l$ is a Chebyshev polynomial of the first kind of order $l$
\citep{stegun65}. %Note that in case 1
%The number of basis functions is
%$N_z$ and $l_\mathrm{max}=2(N_Z-1)$ is the highest polynomial order. 
For a vertical field we only use even Chebyshev polynomials 
to satisfy the symmetry condition at $z=0$, so in case 1 we only
consider $z\geq 0$.

The pseudo-spectral coefficients $\bm{U}_n$  are obtained by demanding
the set of linear equations to be satisfied at $N_z$ collocation
points along the vertical direction. We choose these points as the
extrema of $T_{l_\mathrm{max}}$ plus end points, where
$l_\mathrm{max}$ is the highest polynomial order.

The above procedure discretizes the linear equations to a matrix equation,
\begin{align}\label{matrix_eqn}
\bm{M}\bm{w} = 0, 
\end{align}
where $\bm{M}$ is a $4N_z\times 4 N_z$ matrix representing the $L_{ij}$ 
and upper disk boundary conditions, 
and $\bm{w}$ is a vector storing the pseudo-spectral coefficients. 
Starting with an initial guess for $\sigma$, non-trivial solutions to
Eq. \ref{matrix_eqn} are obtained by varying $\sigma$ using Newton-Raphson iteration 
such that $\mathrm{det}\bm{M}=0$ \citep[details can be found in][]{lin12}.

\subsubsection{Non-dimensionalization}\label{non-dim}
In practice we solve the linearized equations in non-dimensional form,
by defining
%We further write 
\begin{align}
  &z=\hat{z}H,\quad k_x =  \hat{k}_x/H,\quad \sigma = \hat{\sigma}\Omega,
  \quad \delta\bm{v} = \csmid 
  \delta\hat{\bm{v}}, \\ 
  &\delta\bm{B} = B_z\delta\hat{\bm{B}},\quad
  \delta\rho = \rho\hat{W}/\hat{c}_s^2,\quad \delta\Phi =
  \csmid^2\delta\hat{\Phi},  
\end{align} 
where $\hat{c}_s=c_s/\csmid$. We also non-dimensionalize other
background quantities, i.e. $\hat{v}_A=v_A/\csmid$,
$\hat{S}=S/\Omega$, $\hat{\kappa}=\kappa/\Omega$,
$\hat{\Omega}_z=\Omega_z/\Omega$ and $\hat{\eta} = \eta/(H^2\Omega)$.

\subsection{Diagnostics}
We visualize results in terms of dimensionless energy densities. We
define 
\begin{align}
  &E_m \equiv \frac{|\delta \hat{\bm{B}}|^2}{2\beta},\\
  &E_g =
  \frac{\hat{\rho}}{\hat{c}_s^2}\left|\real\left(\hat{W}\dd\hat{\Phi}^*\right)\right|,\\  
  &E_k = \frac{1}{2}\hat{\rho}|\delta\hat{\bm{v}}|^2,\\
  &E_t = \frac{\hat{\rho}|\hat{W}|^2}{2\hat{c}_s^2},
\end{align}
as the perturbed magnetic, gravitational, kinetic and thermal
energies, respectively, which are functions of $z$. The total energy
is $E=E_m+E_g+E_k+E_t$ and we then normalize $E_i$ by
$\mathrm{max}(E)$.

### 2014-02-10 paper/results.tex
major update to tex

\section{Results}

\subsection{MRI in self-gravitating polytropic disks with uniform
  resistivity}  
We first calculate the MRI as a function of $Q$ in a polytropic disk
with uniform resistivity ($A=1$). We consider $Q\in[0.2,4]$,
corresponding to stronly self-gravitating to non-self-gravitating
disks, for mid-plane Elsasser numbers $\Lambda_0\in[0.3,5]$. 
We fix $k_x=0.1$ and $\beta=100$. The numerical resolution is
$N_z=256$.

Fig. \ref{compare_growth_poly_uniresis} plots the MRI growth rates as
a function of $Q$ and $\Lambda_0$. As expected growth rates generally
decrease with $\Lambda_0$. In the limit of ideal MHD, $\Lambda_0>1$,
there is negligible dependence on $Q$. In the resistive limit,
$\Lambda_0<1$, growth rates decrease noticeably for $Q<1$. This shows
show that self-gravity can still affect the MRI even  
though density and potential perturbations are expected to be
negligible because $k_x\ll1$ (i.e. the linear response is
non-self-gravitating).

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth_poly_uniresis}
  \caption{MRI growth rates as a function of $Q$ for mid-planet
    Elsasser numbers $\Lambda_0=0.3$ (solid), $0.5$ (dotted), $1.0$
    (dashed) and $5.0$ (dash-dot).  
    \label{compare_growth_poly_uniresis}}
\end{figure}

% (We find the weak variations in
%$\gamma$ for small $Q$ is associated with a change in the number of
%nodes in the magnetic field perturbation.)
%because $f$ is of order unity for the
%range of $Q$ considered so $\lambda_\mathrm{ideal}$ remain less than
%  unity.
%However, in the resistive limit, the dependence of
%  $\lambda_\mathrm{resis}$ on $f$ (and hence $Q$) is amplified by
%  $\Lambda_0 < 1$.

\cite{sano99} found that for instability to occur the linear mode
dimensional wavelength $\lambda$ should fit inside the disk. That is,   
\begin{align}\label{sano_crit}
  \frac{\lambda}{2H} \equiv
  \mathrm{max}\left(\lambda_\mathrm{ideal},\lambda_\mathrm{resis}\right)\lesssim
  1, 
\end{align}
where the non-dimensional wavelengths are given by 
\begin{align}\label{lambda_ideal}
  \lambda_\mathrm{ideal} = \frac{4\pi}{\sqrt{15}} f v_A =
  \frac{4\pi f}{\sqrt{15\beta\rho}}
\end{align}
in the limit of ideal MHD, and 
\begin{align}\label{lambda_resis}
  \lambda_\mathrm{resis} = \frac{2\pi}{\sqrt{3}}\frac{\eta}{v_A f} =
  \frac{2\pi f}{\Lambda_0}\sqrt{\frac{\rho}{3\beta}} 
\end{align}
%factor pi comes form converting vertical wavenumber to
%wavelength. linear theory works with wavenumber, physical
%interpretation in terms of wavelengths
in the limit of high resistivity. 
%For $\Lambda_0\gtrsim 1$, we find
%the condition $\lambda < 2H$ is well-satisifed throughout most of the
%disk regardless of $Q$.     
For polytropic disks the normalized density $\rho$ is weakly dependent
on $Q$ (Fig. \ref{eqm_den}) so self-gravity only affects the
MRI through the factor $f$, which increases with decreasing $Q$ (see
Fig. \ref{plot_fq} in Appendix \ref{appen1}). This suggests that
sufficiently strong self-gravity can stabilize the MRI by making
$\lambda > 2H$.

%Physically, the
%disk becomes thinner with increasing self-gravity, and the MRI will be
%surpressed when the disk is too thin to accommodate unstable modes. 
%MRI requires
%$\lambda\lesssim 1$, so that the unstable wavelength fits within the
%disk column. As shown in Appendix \ref{appen1}, $f$ increases with the
%strength of self-gravity. This is expected to increase $\lambda$ and
%weaken the MRI.

However, $f$ does not vary much for the range of $Q$ considered here. 
In the ideal limit, we find $\lambda < 2H$ throughout most of the disk
for all values of $Q$, so that self-gravity does not affect growth
rates significantly. Weak variations in $\gamma$ were found to be
associated with changes in the

\begin{figure}
  \includegraphics[width=\linewidth]{figures/lambda_poly_uniresis}
  \caption{Wavelength of the most unstable resistive MRI modes as given by
    Eq. \ref{sano_crit}---\ref{lambda_resis}, normalized by the 
    disk thickness, as a function of height for a range of Toomre $Q$
    values.  
    \label{lambda_poly_resis}}
\end{figure}

\subsection{Resistive MRI in massive polytropic disks with layered
  resistivity}

\subsection{GI in magnetized isothermal disks with uniform
  resistivity}

\subsection{GI in magnetized isothermal disks with layered
  resistivity}

>>>>

\section{Results: MRI in self-gravitating disks}

\subsection{Influence of self-gravity on the MRI through the
  background equilibrium}

In this section we focus on the MRI. We consider polytropic disks,
which have a well-defined thickness, and a vertical field. 
We set the upper disk boundary at $Z_s=0.99H$ and fix $\beta=100$ and
$k_xH=0.1$. The numerical resolution is $N_z=256$.

%kx -> 0 so incompressible

\subsubsection{Uniform
  resistivity}  
We first calculate the MRI as a function of $Q$ with uniform
resistivity ($A=1$) for mid-plane Elsasser numbers $\Lambda_0\in[0.3,10]$.

Fig. \ref{compare_growth_poly_uniresis} plots the MRI growth rates as
a function of $Q$ and $\Lambda_0$. As expected growth rates generally
decrease with increasing $\Lambda_0$. In the limit of ideal MHD ($\Lambda_0>1$), 
there is negligible dependence on $Q$. In the resistive limit
($\Lambda_0<1$), however, growth rates decrease noticeably for $Q<1$. This shows
show that disk self-gravity affects the MRI through the background
equilibrium, even though density and potential perturbations are  
negligible (i.e. the linear response is
non-self-gravitating).

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth_poly_uniresis2}
  \caption{MRI growth rates as a function of $Q$ for mid-planet
    Elsasser numbers $\Lambda_0=0.3$ (solid), $0.5$ (dotted), $1.0$
    (dashed) and $5.0$ (dash-dot).  
    \label{compare_growth_poly_uniresis}}
\end{figure}

\cite{sano99} found that for instability to occur the linear mode
wavelength $\lambda$ should fit inside the disk. That is,   
\begin{align}\label{sano_crit}
  \lambda \equiv
  \mathrm{max}\left(\lambda_\mathrm{ideal},\lambda_\mathrm{resis}\right)\lesssim
  2H, 
\end{align}
where the MRI wavelengths are given by 
\begin{align}\label{lambda_ideal}
  \frac{\lambda_\mathrm{ideal}}{2H} = \frac{4\pi}{\sqrt{15}} f \hat{v}_A =
  \frac{4\pi f}{\sqrt{15\beta\hat{\rho}}}
\end{align}
in the limit of ideal MHD, and 
\begin{align}\label{lambda_resis}
  \frac{\lambda_\mathrm{resis}}{2H} = \frac{2\pi}{\sqrt{3}}\frac{\hat{\eta}}{\hat{v}_A f} =
  \frac{2\pi f}{\Lambda_0}\sqrt{\frac{\hat{\rho}}{3\beta}} 
\end{align}
in the limit of high resistivity.

Since $\hat{\rho}$ is weakly dependent
on $Q$ (Fig. \ref{eqm_den}), self-gravity only affects the
MRI through the factor $f$, which increases with decreasing $Q$ (see
Fig. \ref{plot_fq} in Appendix \ref{appen1}). This suggests that
sufficiently strong self-gravity can stabilize the MRI by making
$ 2H<\lambda $.

In the ideal limit, we find $\lambda < 2H$ throughout most of the disk
for the values of $Q$ considered, so that self-gravity does not affect
growth rates significantly. However, the wavelength of the instability
increases. This is shown in Fig. \ref{compare_result_lambda10} which
plots the magnetic energies for $\Lambda_0=10$ and a range of Toomre $Q$
values. The number of vertical nodes decrease with $Q$, 
i.e. the disk accommodates fewer wavelengths because increasing
vertical self-gravity makes it thinner.

%Weak variations in growth rates in the ideal limit, seen in 
%Fig. \ref{compare_growth_poly_uniresis}, were found to be associated
%with changes in the mode character: $\gamma$ decreases
%slightly when the mode begins to acquire an additional vertical node as $Q$ is
%increased.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_result_lambda10}
  \caption{Magnetic energies as a function of height, in the limit of
    ideal MHD, for various strengths of self-gravity.   
    \label{compare_result_lambda10}}
\end{figure}

Self-gravity appreciably decreases the MRI growth rates in the
resistive limit. Fig. \ref{lambda_poly_resis} plots 
Eq. \ref{sano_crit} for $\Lambda_0=0.3$. In the non-self-gravitating
disk ($Q=4$) the instability criterion is marginally satisfied and the
MRI operates. As $Q$ decreases, 
Eq. \ref{sano_crit} is violated and the MRI growth rate is
significantly reduced. This is seen for $Q=0.2$ where $\lambda \geq 2H$ throughout
the disk. (The instability is not completely surpressed since
Eq. \ref{sano_crit} is only an approximation.) Although the function
$f$ does not change significantly for the range of $Q$ considered, the
dependence of $\lambda_\mathrm{resis}$ on $f$ (and therefore $Q$) is
amplified by the denominator $\Lambda_0<1$ in the resistive case. For
resistive cases we find no nodes in the magnetic energy $E_1$ except
at end points, i.e. only the longest wavelength survives.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/lambda_poly_uniresis}
  \caption{Approximate wavelengths of the most unstable resistive MRI modes as given by
    Eq. \ref{sano_crit}---\ref{lambda_resis}, normalized by the 
    disk thickness, as a function of height for a range of Toomre $Q$
    values.  
    \label{lambda_poly_resis}}
\end{figure}

\subsubsection{Layered
  resistivity} 
Here we consider disks with mid-plane Elsasser number $\Lambda_0=0.1$
and a variable resistivity profile with
$A=10^2$. Fig. \ref{poly_layer} compares the magnetic  
energies for $Q=0.2,\,1$ and $4$. They have similar growth rates, $\gamma/\Omega
= 0.53,\,0.64$ and $0.66$, respectively. In the non-self-gravitating
limit ($Q=4$), the MRI is effectively surpressed for
$z\lesssim0.5H$. This is consistent with layered accretion. However,
in the massive disk ($Q=0.2$) the mode occupies a wider vertical
extent because its wavelength (in units of $H$) is larger. This
suggests that in massive disks the MRI is not well localized to the
active layer, as the instability takes on a more global character
compared to non-self-gravitating disks.        
%d is not as well localized

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_results_poly_layer_amp100}
  \caption{Magnetic energies as a function of height, for layered polytropic disks with
    such that the conductivity increases by a 
    factor of $10^2$ in going from the mid-plane to the upper disk
    boundary. The vertical lines indicate $\Lambda=1$ for each value
    of $Q$.
    \label{poly_layer}}
\end{figure}

\subsection{Influence of self-gravity on the MRI through the linear
  response}

Our goal here is to examine whether or not self-gravity can amplify
the density perturbations caused by the MRI. We choose a massive 
isothermal disk with $Q=0.2$, which is still expected to be marginally
stable to gravitational instability \citep[][who find
a critical value of  $Q\simeq 0.2$]{mamat10}.  
The upper disk boundary is set to $Z_s=H$. The field is vertical.

%resolution Nz=256

\subsubsection{Ideal case}
We first consider the limit of ideal MHD by adopting a uniform
resistivity with $\Lambda_0=100$. %Non-zero density perturbations
%require $k_x\neq0$ so we consider linear modes as a function of $k_x$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_energy_ideal}
  \caption{Maximum gravitational potential energy perturbation
    associated with the MRI in a self-gravitating disk with $Q=0.2$ in
    the limit of ideal MHD ($\Lambda_0=10^2$). 
    \label{gravity_energy}}
\end{figure}

\subsubsection{Qualitative interpretation} 
In order to make sense of the above results, we consider ideal MHD and
a vertical field. For this discussion we will ignore stratification
and set $d/dz\to \imgi k_z$ where $k_z$ is a vertical 
wavenumber. Equivalently, we are considering regions close to the disk
midplane. The governing equations are then

\begin{align}
  &  0= v_A^2k^2\dvx + \imgi\sigma\left(\imgi\sigma \dvx - 2\Omega\dvy + \imgi k_x \w\right),\label{simp1}\\
  &  0= v_A^2k_z^2\left(\dvy + \frac{\imgi S
  }{\sigma}\dvx\right) + \imgi\sigma\left(\imgi\sigma\dvy +
  \frac{\kappa^2}{2\Omega}\dvx\right),\\
  & 0 = -k_z^2\w + \frac{\sigma^2}{c_s^2}W + \sigma k_x \dvx, \label{simp3}\\
  & 0 = k^2\dphi + \frac{\Omega^2}{c_s^2Q}W,\label{simp4}
\end{align}
where $k^2 = k_z^2 + k_x^2$.

We imagine an iterative procedure to solve the above equations,
starting from the \emph{Cowling approximation}, defined to be the
solution to Eq. \ref{simp1}---\ref{simp3} with potential perturbation
set to zero. This is the standard MRI and we denote the solution as
$\dvx^{(0)}$, $\dvy^{(0)}$ and $W^{(0)}$.

We now include self-gravity. In general, the MRI will have some
density perturbation so $W^{(0)}\neq0$. The Poisson equation implies
an associated potential perturbation,    
\begin{align} 
  \dphi = -\frac{\Omega^2}{c_s^2 Q k^2} W^{(0)}.
\end{align}
Physically, we expect $k^2\geq0$, so that a positive (negative) density
perturbation causes a negative (positive) potential perturbation. We
then regard $\dphi$ to alter the Cowling solution, so
that $\dvx^{0} \to \dvx^{0} + \dvx^{(1)}$ and similarly for $\dvy$ and
$W$. Then 
\begin{align}
  &   k_x\sigma \dphi = v_A^2k^2\dvx^{(1)} + \imgi\sigma\left[\imgi\sigma
  \dvx^{(1)} - 2\Omega\dvy^{(1)} + \imgi k_x W^{(1)}\right], \label{simp_pert1}\\ 
  &  0= v_A^2k_z^2\left[\dvy^{(1)} + \frac{\imgi S
    }{\sigma}\dvx^{(1)}\right] + \imgi\sigma\left[\imgi\sigma\dvy^{(1)} +
  \frac{\kappa^2}{2\Omega}\dvx^{(1)}\right],\label{simp_pert2}\\
  & k_z^2\dphi  = \left(\frac{\sigma^2}{c_s^2}-k_z^2\right)W^{(1)} +
  \sigma k_x \dvx^{(1)}.\label{simp_pert3} 
\end{align}
Now, if the perturbations to the magnetic field remain
unchanged, i.e. the mode remains close to the standard MRI, then
$\dvx^{(1)} \sim 0$ and $\dvy^{(1)}\sim0$, so Eq. \ref{simp_pert2} is
satisfied. Eq. \ref{simp_pert1} then require $\dphi +
W^{(1)}\sim0$. This is compatible with Eq. \ref{simp_pert3} if 
\begin{align}
  \left|k_z^2\right| \gg \left|\frac{\sigma^2}{c_s^2}\right|. \label{cond}
\end{align}
These assumptions imply
\begin{align}
  W^{(1)} \sim \frac{\Omega^2}{c_s^2 Q k^2} W^{(0)},\label{feedback}
\end{align} 
which indicates a non-zero density perturbation due to the
MRI can be amplified by self-gravity. 
%However, the magnetic field
%perturbations remain that in the

By introducing $\hat{k}_z=k_zH$, Eq. \ref{cond} can be
non-dimensionalized to 
\begin{align}
  \left|\frac{\hat{\sigma}^2}{f^2\hat{c}_s^2\hat{k}_z^2}\right|\ll 1.  
\end{align}

%\begin{align}
%  
%\end{align}
%However, in the full problem $\dphi$ appears in the $\mathcal{G}$

%We write the horizontal velocities and enthalpy perturbations as
%\begin{align}
%  \dvx &= u + \Delta u,\\
%  \dvy &= v + \Delta v,\\
%  W    &= w + \Delta w,
%\end{align}
%where $u$, $v$ and $w$ is the solution to
%Eq. \ref{simp1}---\ref{simp4} in the Cowling approximation, in which
%the Poisson equation is ignored and the potential perturbation set to
%zero.
